---
title: "Housing Price Model"
author: "Anurag Nepal"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    theme: null
    css: blog.css
    includes:
      before_body: nav.html
      after_body: footer.html
  word_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.align = "center")

library(tidyverse)
library(caret)
library(glmnet)
library(corrplot)
library(gridExtra)
library(naniar)
library(dplyr)

set.seed(748)
```

# Background

I wanted to know which preprocessing choices actually help when modeling house prices. The usual advice is to impute missing values, normalize features, cap outliers, and log-transform skewed targets. Instead of taking that on faith, I ran a bake-off: five pipelines from a plain complete-case baseline up to full preprocessing, each tested with OLS, Ridge, and Lasso under 10-fold cross-validation.

# Setup

The data is `housing_data.csv`, in this repo. Each row is a house with size in square feet, bedroom count, age, distance to the city, crime rate, and renovation status, plus the sale price. I split 80/20 into train and test right at the start, before any exploration, so nothing from the test set could leak into the EDA or the preprocessing.

## Data provenance

This dataset came to me without documentation of where the records were collected, and I have not been able to trace a source. Several values look generated rather than measured: fractional square footage (1703.96 sqft), house ages with sub-day precision, and prices to the cent with no heaping. I treat it as a synthetic dataset and read the results accordingly. The bake-off still compares preprocessing choices fairly on this data, but the absolute error levels should not be quoted as real-world performance. If I trace the source later, I will cite it here.

```{r load_data}
raw_housing_df <- read.csv("housing_data.csv")

raw_housing_df$renovated <- as.factor(raw_housing_df$renovated)

split_indices <- createDataPartition(raw_housing_df$price, p = 0.8, list = FALSE)
housing_train <- raw_housing_df[split_indices, ]
housing_test <- raw_housing_df[-split_indices, ]
```

# Method

## Exploratory analysis

I started with the structure of the training data and where the missing values sit.

```{r eda_structure}
summary(housing_train)

na_summary <- data.frame(Missing_Count = colSums(is.na(housing_train)))
knitr::kable(na_summary, caption = "Missing Values by Feature")
```

Missing values show up in a few predictors (`size_sqft`, `age`, `price_per_sqft`). Dropping every incomplete row would throw away a big chunk of training data, so imputation is on the table. Before deciding anything, I checked whether the missingness looked random.

```{r mcar-test}
mcar_result <- mcar_test(raw_housing_df)
print(mcar_result)
```

Little's MCAR test came back with p = 0.223, so I cannot reject the idea that values are missing completely at random. That matters later: if missingness carries no signal, dropping rows costs little and imputing adds nothing.

Next I looked at distributions. Both price and crime rate are heavily right-skewed, with a long tail of expensive homes and high-crime areas. Distance to the city has clear outliers on the boxplot.

```{r eda_distributions, fig.width=10, fig.height=8}
plot_price_dist <- ggplot(housing_train, aes(x = price)) + 
  geom_histogram(bins = 30, fill = "steelblue", color = "black") + 
  theme_minimal() + 
  labs(title = "Price Distribution (Skewed)", x = "Price", y = "Count")

plot_crime_dist <- ggplot(housing_train, aes(x = crime_rate)) + 
  geom_histogram(bins = 30, fill = "coral", color = "black") + 
  theme_minimal() + 
  labs(title = "Crime Rate Distribution", x = "Crime Rate", y = "Count")

plot_dist_outliers <- ggplot(housing_train, aes(y = distance_city)) + 
  geom_boxplot(fill = "lightgreen") + 
  theme_minimal() + 
  labs(title = "Distance to City Outliers", y = "Distance")

grid.arrange(plot_price_dist, plot_crime_dist, plot_dist_outliers, ncol=2)
```

The correlation check turned up two things worth acting on. First, `price_per_sqft` correlates almost perfectly with price, because it is computed from price and size. That is target leakage, so it has to go. Second, `size_sqft` and `bedrooms` correlate strongly, which is expected: more bedrooms need more area.

```{r eda_correlation, fig.width=8, fig.height=6}
numeric_subset <- housing_train %>% dplyr::select_if(is.numeric)
correlation_matrix <- cor(numeric_subset, use = "complete.obs")

corrplot(correlation_matrix, method = "color", type = "upper", 
         addCoef.col = "black", tl.col = "black", diag = FALSE)
```

That gave me the preprocessing plan:

1. **Drop the leakage.** Remove `price_per_sqft`.
2. **Handle missing values.** Median for numbers, mode for categories, plus missingness indicators in one pipeline to test whether the missingness itself carries signal.
3. **Handle outliers.** Cap at the 5th and 95th percentiles rather than deleting rows.
4. **Transform and scale.** Standardize predictors for Ridge and Lasso, and log-transform the skewed `price` and `crime_rate`.

## The five pipelines

Every pipeline below is trained with 10-fold cross-validation and scored on the held-out test set. Two helpers keep the bookkeeping straight: one pulls the best CV row out of each caret fit, and one scores the test set. The imputation helper always uses training-set statistics, so the test set never leaks into training.

```{r cv_setup}
# Cross-Validation Configuration
cv_control <- trainControl(method = "cv", number = 10)

# Summary function for Log-transformed models (Back-transforms to original scale)
exp_summary <- function(data, lev = NULL, model = NULL) {
  obs <- exp(data$obs)
  pred <- exp(data$pred)
  c(RMSE = RMSE(pred, obs), Rsquared = R2(pred, obs), MAE = MAE(pred, obs))
}

cv_control_log <- trainControl(method = "cv", number = 10, summaryFunction = exp_summary)

# Storage for results
performance_metrics_storage <- list()
test_metrics_storage <- list()
```


```{r cv_setup_test}
# Extract CV performance from caret object
extract_performance <- function(model_obj, pipeline_label, model_type) {
  best_row <- model_obj$results[which.min(model_obj$results$RMSE), ]
  data.frame(
    Pipeline = pipeline_label,
    Model = model_type,
    RMSE = paste0(round(best_row$RMSE, 2), " (± ", round(best_row$RMSESD, 2), ")"), 
    MAE = paste0(round(best_row$MAE, 2), " (± ", round(best_row$MAESD, 2), ")"),
    RMSE_Num = best_row$RMSE
  )
}

# Calculate performance on unseen test set
get_test_perf <- function(model_obj, test_data, label, model_type, is_log = FALSE) {
  preds <- predict(model_obj, newdata = test_data)
  obs <- test_data$price
  if(is_log) preds <- exp(preds)
  
  data.frame(
    Pipeline = label,
    Model = model_type,
    Test_RMSE = round(RMSE(preds, obs), 2),
    Test_MAE = round(MAE(preds, obs), 2)
  )
}

# Unified Imputation Function (Ensures Test uses Training stats)
impute_data <- function(target_df, reference_df) {
  res_df <- target_df
  for(nm in names(res_df)) {
    if(is.numeric(res_df[[nm]])) {
      res_df[[nm]][is.na(res_df[[nm]])] <- median(reference_df[[nm]], na.rm = TRUE)
    } else {
      mode_val <- names(sort(table(reference_df[[nm]]), decreasing = TRUE))[1]
      res_df[[nm]][is.na(res_df[[nm]])] <- mode_val
    }
  }
  return(res_df)
}
```

### Pipeline 0: baseline

Drop the leakage variable and any row with a missing value. No scaling, no transformations. This is the complete-case OLS baseline everything else has to beat.

```{r pipeline_0}
housing_train_clean <- housing_train %>% dplyr::select(-price_per_sqft)
housing_test_clean  <- housing_test %>% dplyr::select(-price_per_sqft)

train_p0 <- na.omit(housing_train_clean)
test_p0  <- na.omit(housing_test_clean)

# Train
fit_p0_ols   <- train(price ~ ., data = train_p0, method = "lm", trControl = cv_control)
fit_p0_ridge <- train(price ~ ., data = train_p0, method = "glmnet", 
                      tuneGrid = expand.grid(alpha = 0, lambda = seq(0.1, 10, length=5)),
                      trControl = cv_control)

performance_metrics_storage[[1]] <- extract_performance(fit_p0_ols, "P0_Baseline", "OLS")
performance_metrics_storage[[2]] <- extract_performance(fit_p0_ridge, "P0_Baseline", "Ridge")
test_metrics_storage[[1]] <- get_test_perf(fit_p0_ols, test_p0, "P0_Baseline", "OLS")
test_metrics_storage[[2]] <- get_test_perf(fit_p0_ridge, test_p0, "P0_Baseline", "Ridge")
```

### Pipeline 1: imputation

Keep all rows by filling missing numbers with the median and missing categories with the mode. Still no scaling.

```{r pipeline_1}
train_p1 <- impute_data(housing_train_clean, housing_train_clean)
test_p1  <- impute_data(housing_test_clean, housing_train_clean)

# Train
fit_p1_ols   <- train(price ~ ., data = train_p1, method = "lm", trControl = cv_control)
fit_p1_ridge <- train(price ~ ., data = train_p1, method = "glmnet", 
                      tuneGrid = expand.grid(alpha = 0, lambda = seq(0.1, 10, length=5)),
                      trControl = cv_control)

# Evaluate
performance_metrics_storage[[3]] <- extract_performance(fit_p1_ols, "P1_Imputed", "OLS")
performance_metrics_storage[[4]] <- extract_performance(fit_p1_ridge, "P1_Imputed", "Ridge")
test_metrics_storage[[3]] <- get_test_perf(fit_p1_ols, test_p1, "P1_Imputed", "OLS")
test_metrics_storage[[4]] <- get_test_perf(fit_p1_ridge, test_p1, "P1_Imputed", "Ridge")
```


### Pipeline 2: imputation plus missingness indicators

Same as Pipeline 1, but with a binary column flagging where each missing value was before imputation. If missingness carries signal about price, this should pick it up. Near-zero-variance columns are dropped.

```{r pipeline_2}
# Add Indicators
add_indicators <- function(df) mutate(df, across(everything(), is.na, .names = "na_{.col}"))

train_p2_raw <- add_indicators(housing_train_clean)
test_p2_raw  <- add_indicators(housing_test_clean)

# 2. Impute
train_p2_imputed <- impute_data(train_p2_raw, train_p2_raw)
test_p2_imputed  <- impute_data(test_p2_raw, train_p2_raw)

# This identifies columns where all values are the same (e.g., all FALSE)
nzv_cols <- nearZeroVar(train_p2_imputed)

if(length(nzv_cols) > 0) {
  train_p2 <- train_p2_imputed[, -nzv_cols]
  test_p2  <- test_p2_imputed[, -nzv_cols]
} else {
  train_p2 <- train_p2_imputed
  test_p2  <- test_p2_imputed
}

# Train
fit_p2_ols   <- train(price ~ ., data = train_p2, method = "lm", trControl = cv_control)
fit_p2_ridge <- train(price ~ ., data = train_p2, method = "glmnet", 
                      tuneGrid = expand.grid(alpha = 0, lambda = seq(0.1, 10, length=5)),
                      trControl = cv_control)

# Evaluate
performance_metrics_storage[[5]] <- extract_performance(fit_p2_ols, "P2_Indicators", "OLS")
performance_metrics_storage[[6]] <- extract_performance(fit_p2_ridge, "P2_Indicators", "Ridge")
test_metrics_storage[[5]] <- get_test_perf(fit_p2_ols, test_p2, "P2_Indicators", "OLS")
test_metrics_storage[[6]] <- get_test_perf(fit_p2_ridge, test_p2, "P2_Indicators", "Ridge")
```

### Pipeline 3: normalization

Standardize the numeric predictors using training-set means and standard deviations, then fit Ridge and Lasso. Regularized models need this so the penalty treats features fairly.

```{r pipeline_3}
# Identify numeric columns (excluding price)
num_cols <- names(train_p2)[sapply(train_p2, is.numeric) & names(train_p2) != "price"]

# Calculate scaling parameters from train
train_means <- colMeans(train_p2[, num_cols])
train_sds   <- apply(train_p2[, num_cols], 2, sd)

# Scale both
train_p3 <- train_p2
test_p3  <- test_p2
train_p3[num_cols] <- scale(train_p2[num_cols], center = train_means, scale = train_sds)
test_p3[num_cols]  <- scale(test_p2[num_cols], center = train_means, scale = train_sds)

# Train
fit_p3_ridge <- train(price ~ ., data = train_p3, method = "glmnet", 
                      tuneGrid = expand.grid(alpha = 0, lambda = seq(0.1, 10, length=5)),
                      trControl = cv_control)
fit_p3_lasso <- train(price ~ ., data = train_p3, method = "glmnet", 
                      tuneGrid = expand.grid(alpha = 1, lambda = seq(0.1, 10, length=5)),
                      trControl = cv_control)

# Evaluate
performance_metrics_storage[[7]] <- extract_performance(fit_p3_ridge, "P3_Normalized", "Ridge")
performance_metrics_storage[[8]] <- extract_performance(fit_p3_lasso, "P3_Normalized", "Lasso")
test_metrics_storage[[7]] <- get_test_perf(fit_p3_ridge, test_p3, "P3_Normalized", "Ridge")
test_metrics_storage[[8]] <- get_test_perf(fit_p3_lasso, test_p3, "P3_Normalized", "Lasso")
```

### Pipeline 4: full preprocessing

Everything at once: imputation, indicators, log-transforms of price and `crime_rate`, Winsorization at the 5th and 95th percentiles using training quantiles, then re-standardization. The log-scale models are scored back on the original dollar scale so the comparison stays fair.

```{r pipeline_4}
# Log Transforms
train_p4 <- train_p2 %>% mutate(price = log(price), crime_rate = log(crime_rate + 1))
test_p4  <- test_p2  %>% mutate(crime_rate = log(crime_rate + 1)) # Price stays original for get_test_perf

# Capping (Winsorization) using Training Quantiles
for(nm in num_cols) {
  qnts <- quantile(train_p4[[nm]], probs = c(.05, .95), na.rm = TRUE)
  train_p4[[nm]] <- pmin(pmax(train_p4[[nm]], qnts[1]), qnts[2])
  test_p4[[nm]]  <- pmin(pmax(test_p4[[nm]], qnts[1]), qnts[2])
}

# Re-scale based on the new P4 training distributions
train_p4_means <- colMeans(train_p4[, num_cols])
train_p4_sds   <- apply(train_p4[, num_cols], 2, sd)

train_p4[num_cols] <- scale(train_p4[num_cols], center = train_p4_means, scale = train_p4_sds)
test_p4[num_cols]  <- scale(test_p4[num_cols], center = train_p4_means, scale = train_p4_sds)

# Train
fit_p4_ridge <- train(price ~ ., data = train_p4, method = "glmnet",
                      tuneGrid = expand.grid(alpha = 0, lambda = seq(0.001, 0.1, length=5)),
                      trControl = cv_control_log)
fit_p4_lasso <- train(price ~ ., data = train_p4, method = "glmnet",
                      tuneGrid = expand.grid(alpha = 1, lambda = seq(0.001, 0.1, length=5)),
                      trControl = cv_control_log)

# Evaluate
performance_metrics_storage[[9]] <- extract_performance(fit_p4_ridge, "P4_Full_Processing", "Ridge")
performance_metrics_storage[[10]] <- extract_performance(fit_p4_lasso, "P4_Full_Processing", "Lasso")

test_metrics_storage[[9]] <- get_test_perf(fit_p4_ridge, test_p4, "P4_Full_Processing", "Ridge", is_log = TRUE)
test_metrics_storage[[10]] <- get_test_perf(fit_p4_lasso, test_p4, "P4_Full_Processing", "Lasso", is_log = TRUE)

```

# Results

The cross-validated comparison and the held-out test scores tell the same story.

```{r comparison_table}
final_results <- bind_rows(performance_metrics_storage)
knitr::kable(final_results %>% dplyr::select(-RMSE_Num), caption = "Final Model Comparison")
```

```{r comparison_table_test}
test_results_final <- bind_rows(test_metrics_storage)
knitr::kable(test_results_final, caption = "Model Performance on Unseen Test Data")
```

```{r performance_plot, fig.width=9, fig.height=6}
ggplot(final_results, aes(x = reorder(Pipeline, RMSE_Num), y = RMSE_Num, fill = Model)) +
  geom_bar(stat = "identity", position="dodge") +
  coord_flip() + theme_minimal() + labs(title = "RMSE by Pipeline", y = "RMSE (Original Scale)")
```

The headline results:

* **The plain baseline won.** Pipeline 0 OLS had the lowest CV RMSE (45,814.87) and MAE (31,707.57), and stayed best on the test set (RMSE 63,202.49, MAE 36,907.90). Every fancier pipeline did worse.
* **Complete-case analysis did not hurt.** Dropping incomplete rows beat all imputation. That makes sense given the MCAR test: with missingness completely at random, dropping rows leaves a smaller but unbiased sample, while median/mode imputation compresses variance and adds noise. CV RMSE rose from ~46k (P0) to ~51k (P1).
* **Missingness indicators added nothing.** Going from P1 to P2 pushed OLS RMSE from ~51k to ~53k. The indicators carried no signal about price, which is exactly what MCAR predicts.
* **Normalization did not save the regularized models.** It helped Lasso slightly relative to the imputation pipelines, but everything stayed worse than the baseline. In housing, large-scale features like square footage carry the strongest signal, and standardizing mutes that natural importance while the penalty bites harder.
* **Winsorizing hurt the most.** Pipeline 4 came in worst overall at ~58k RMSE. The extremes here are informative: genuine luxury homes with genuinely high prices. Capping them forced systematic under-prediction at the top end.
* **The log transform backfired on the dollar scale.** Back-transforming small log-scale errors into dollars blew them up, pushing MAE to ~39k. The price relationship in this data looks additive, not multiplicative.

### Pipeline 4b: smearing-corrected log back-transform (back-transformation sensitivity analysis)

The Pipeline 4 models are fit on `log(price)` and scored back on the dollar scale with a plain `exp()`. By Jensen's inequality that back-transform estimates `exp(E[log Y | x])` rather than `E[Y | x]`, so it systematically under-predicts and can inflate the log-pipeline's dollar RMSE relative to the other pipelines. As a follow-up sensitivity check, `analysis/smearing_sensitivity.R` (run: `Rscript analysis/smearing_sensitivity.R` from the repo root) re-runs all five pipelines on a fresh seeded 80/20 split and evaluates the P4 log models under three back-transformations on the same fits: the naive `exp()`, Duan's smearing correction `exp(pred) * mean(exp(residuals))` (Duan, 1983), and the parametric lognormal correction `exp(pred) * exp(sigma^2/2)`. The original naive-exp comparison above is untouched; this is the same data and the same fits with only the back-transform varied.

```{r smearing-table}
smear_comp = read.csv("analysis/smearing_comparison.csv")
knitr::kable(smear_comp, caption = "Test performance: all pipelines, with P4 under three back-transforms")
```

The smearing factors are small, 1.0258 for Ridge and 1.0255 for Lasso, and the correction moves the P4 test RMSE slightly the wrong way: Ridge goes from 59,362 (naive) to 59,881 (smearing), Lasso from 62,106 to 63,129. The parametric correction lands in the same place. So the sensitivity analysis confirms the original ranking rather than overturning it: the log transform's poor showing is not an artifact of the naive back-transform, and Pipeline 4 stays last either way. Note this re-run uses a fresh seeded split (the original analysis did not seed the split), so absolute numbers here differ slightly from the tables above; what matters is the within-run comparison, where data and fits are held fixed.

# Takeaway

Preprocessing is not free, and on this data most of it was harmful. Because the missingness was completely at random, the simplest thing, dropping incomplete rows and fitting OLS, beat every imputation, scaling, and transformation I tried. The two mistakes to avoid generalize: keeping a leakage variable like `price_per_sqft` would let the model cheat off the target's own formula, and computing imputations or scaling factors before the train/test split would leak test information into training. Both inflate performance estimates that do not survive contact with new data.
